---
title: "R for Biochemists 201, Module 2, Calculating and adding p-values"
author: "Paul Brennan, Cardiff University, BrennanP@cardiff.ac.uk"
date: "Nov 2021"
output: 
#   html_document
   word_document: default
#  pdf_document: default
---

### Welcome to module 2 for Biochemists 201 Training module.

In Module 1, we have consolidated some of our R skills, gained an
understanding of tidy data and learnt how to create some more complex plots.
In Module 2, we will look at calculating some simple statistics and adding
the statistics to a multipanel plot. Again we seek to revisit and deepen
the skills from R for Biochemists 101. 

Our exercises are more demanding so please allow a little more time
to complete the material. In our pilot, learners took a little over 2 hours
to complete this module. Please don't worry if it takes you longer.
Rather be patient with yourself and try to keep learning. Take a break as 
you need and ask questions if you need help. 

As before, the material provided includes text to read, video demonstrations,
example R scripts with exercises. We hope you enjoy the module.

We welcome feedback and comments.

### How to use this module 
1. To learn from this module, cut and paste the R code into R-Studio and 
then run the code line by line. 
2. See if you can make the script work
3. Watch the demonstration videos to see how the code works
4. Look at how the 'Global Environment' changes.
5. Change the code and test it
6. Try the exercises at the end
7. Extend your learning using the resources at the end

### Learning Objectives for Module 2
* Revisit our data science workflow
* Compare means with T-tests in R
* Add statistics to plots with the ggpubr package
* Prepare a multi-panel figure with R

### Loading the packages we need
```{r packages_we_need}
library(tidyverse)
library(readxl)
library(ggpubr)
```

pagebreak \pagebreak

### The Experiment
We are going to look at a different figure from the Nature paper by Croft et al
available here: [Online](https://www.nature.com/articles/s41586-019-1263-7) and as a [PDF](https://www.nature.com/articles/s41586-019-1263-7.pdf). Also 
in [Biorxiv as preprint](https://www.biorxiv.org/content/10.1101/374330v1). 

We are going to explore the data in [Figure 4a](https://www.nature.com/articles/s41586-019-1263-7/figures/4). The data is [available](https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-019-1263-7/MediaObjects/41586_2019_1263_MOESM8_ESM.xlsx). 
As before the Excel file has 9 worksheets - each sheet corresponding to a 
part of the figure. The authors have drawn 12 bar plots, calculated some
statistics and labelled the plots. In this module, I would like to reproduce
this data.  

As always, the steps are:

1. Import the data
2. Tidy the data
3. Summarise
4. Visualise



### Importing and tidying our data
##### First video - importing and tidying our data 
```{r get_data_for_Fig_4a}
link <- "https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-019-1263-7/MediaObjects/41586_2019_1263_MOESM10_ESM.xlsx"

# the download.file() function downloads and saves the file with the name given
download.file(url=link, destfile="file.xlsx", mode="wb")

# check worksheet names 
sheet_names <- excel_sheets("file.xlsx")
data <- read_excel("file.xlsx", sheet_names[1])
```

So if you look at the data... it's a bit of a challenge. There is lots of
missing
values in the form of NAs. The numbers appear to be on different rows to the
names. All in all it is a real challenge. 

The only way to work out what is going on is to open the file in in Excel. 
If you look at the file in Excel, you notice that the names are offset by
a line.
You can also see that the first six values are one set, there is a gap and
the 
second set of values are a different set. The set names are included in every
row. A little tidying in Excel is the easiest thing to do first. 
all in R. That's what I'm going to do as you shall see in the video.

Now, save this file so that we have a cleaner import..
```{r download_cleaner_data_for_Fig_4a}
link <- ("https://raw.githubusercontent.com/brennanpincardiff/RforBiochemists/master/data/figure_4a_data_altered.xlsx")
download.file(url=link, destfile="file.xlsx", mode="wb")

data_c <- read_excel("file.xlsx")

```

Have a look at the data and it looks much easier to use. It's still
not in a tidy format but it's easier to understand visually. 

We can do all the tidying in one piped together script.
To illustrate each step, we can select look at the output for each step. 
Let's try to tidy this whole data set with the pivot_longer() function.
Then we need to clean up the thy_status which is currently in the column name
and includes numbers and mentions Thy_1. The separate() function works well.
The plotting will be much easier if thy_status and Cytokine are factors which
means they are categories rather than characters. For Cytokine, we put these
in the order that they are plotted in the paper. 
We push this all into a new object called tidy_data. 


```{r tidy_data_for_Fig_4a}
# here we have a selection of functions piped together to form a 
# script to tidy our data...
data_c %>%
    pivot_longer(-Cytokine) %>%
    separate(name, into=c(NA, "thy_status", NA)) %>%
    mutate(thy_status = factor(thy_status)) %>%
    mutate(Cytokine = str_to_upper(Cytokine)) %>%
    mutate(Cytokine = factor(Cytokine, c("CXCL14", "CXCL5", "CXCL2", "CXCL12",
                                         "CCL2", "CCL7", "CCL5", "CCL12","IL6",
                                         "MMP3", "MMP9", "MMP13"))) -> tidy_data


# functions used in this...
# pivot_longer()
    # the pivot_longer() function turns the data into long format. 
    # when we exclude the Cytokine column, the function recognises this
# separate()
    # the separate() function cleans up the thy status column 
    # it uses the _ and . as separators and only keeps the middle values 
    # "plus" and "minus"
# mutate()
    # the mutate() function to create factors using the factor() function
# str_to_upper
    # make the cytokine name all uppercase

```


### Exercise 1
Let's go back to Croft's paper and look at Figure 2d. This is another
multipanel figure. The figure looks like this:
[insert figure, check copyright]
In this case there are six panels. 
The data for Figure 2d is available to download. 

Try the following steps:

1. downloading the data and looking at in Excel. 
2. reogranise in Excel - it'll be easier.
3. save your file and import it into R.
4. then tidy in R.

Please don't try to plot it yet, that's a later exercise...

### Answers for Exercise 1
How did you get on?

My code is available on [Github here](https://github.com/brennanpincardiff/R4Biochemists201/blob/master/R/module_201_2_ans_ex_1.R) as is the Excel file I created when I 
cleaned it up a little. 

You may have cleaned it up differently and so your code may be different. It is 
fine. The key is to progress with the task. 

```{r exercise_1}
# import a cleaner file from Excel
# read in easier data than the download.
link <- ("https://raw.githubusercontent.com/brennanpincardiff/RforBiochemists/master/data/fig_2d_altered.xlsx")
download.file(url=link, destfile="fig2d_altered.xlsx", mode="wb")

data <- read_excel("fig2d_altered.xlsx", skip = 1)

# tidy it with various functions...
# gather(), separate() and mutate() into factors...
tidy_data_2 <- data %>%
    gather() %>%
    separate(key, into=c("names", "DTR_status"), sep = "/") %>%
    # need to turn the names into factors...
    mutate(DTR_status = factor(DTR_status)) %>%
    mutate(names = factor(names, c("bone_eros_CT", "bone_form_CT", 
                                   "area_bone_eros", "area_bone_form",
                                   "area_pan", "area_dest_cart")))


```



### Making a bar chart with T-test
##### Second video - calculating and adding T-tests 

Our next step is to do some statistics. 
We will select one of the charts. Our first chart is CXCL14, a cytokine. 
Let's pull of the CXCL14 data and make
a plot. We use the filter() function to filter rows and create an object 
called cxcl14
```{r start_with_cxcl14}
# start with out data
tidy_data %>%
    # filter the row with the Cxcl14 measurements
    filter(Cytokine == "CXCL14") -> cxcl14

plot <- ggplot(cxcl14,aes(x=thy_status, y= value)) + 
        geom_bar(stat = "summary", fun = "mean")

plot
```


Now let's add a T-test. 
So what is the T-test - theory video here.




```{r calc_t_test_and_add_with_ggpubr}
# calculate stats using a t.test using t.test()
output <- t.test(cxcl14[1:6,3], cxcl14[7:12,3])
output # gives the results
output$p.value # gives the p-value

# the package ggpubr extends ggplot
# we can use the function stat_compare_means with the t.test method. 
# gives us a nice label on our graph...
# Welch Two Sample t-test
plot + stat_compare_means(method = "t.test")
```




pagebreak \pagebreak

### Exercise 2
Repeat this whole process for the cytokine called MMP9: filter, draw a plot, 
calculate a T-test and annotate the graph.
See if you can improve your plot with labels and a source. 

### Answers for Exercise 2

This code is also on [Github]([Github here](https://github.com/brennanpincardiff/R4Biochemists201/blob/master/R/module_201_2_ans_ex_2.R))

```{r exercise_2}
# start with our data
tidy_data %>%
    # filter the row with the Cxcl14 measurements
    filter(Cytokine == "MMP9") -> mmp9

ggplot(mmp9, aes(x=thy_status, y= value)) + 
        geom_bar(stat = "summary", fun = "mean") -> plot

plot

# calculate stats using a t.test using t.test()
output <- t.test(mmp9[1:6,3], mmp9[7:12,3])
output # gives the results
output$p.value # gives the p-value

# the package ggpubr extends ggplot
# we can use the function stat_compare_means with the t.test method. 
# gives us a nice label on our graph...
# Welch Two Sample t-test
plot + stat_compare_means(method = "t.test")

plot + stat_compare_means(method = "t.test") +
    labs(x = "Thy Status",
        y = "MMP9 levels (ng/ml-1)", 
         title = "from Figure 4a",
        subtitle = "Croft et al, Nature, 2019") +
    theme_classic()

```

### Plotting all the data
##### Third video

So the next stage is to plot all the data. 
As we know from module one there are various ways to do this. While I don't 
like bar charts we will continue to use them becuase they are common in 
biochemistry and used in the paper.


```{r summarise_and_bar_chart}
# DATA SUMMARISE
# calculate means and create a new file with the means
tidy_data %>%
    group_by(Cytokine, thy_status) %>%
    summarise(mean =mean(value)) -> cyto_means

# FIRST VISUALISATION of the means
ggplot(cyto_means, aes(x = thy_status, 
    y = mean)) + 
    geom_col() + 
    facet_wrap(~Cytokine, scales ="free_y")
```

Let's make a more complex visualisation. We go back to the data and use 
stat_summary() to calculate the means. Facet_wrap() that we met in R101 works 
well and geom_jitter to add the points. 

```{r more_complex_viz}
# MORE COMPLEX VISUALISATION
# let's try adding the points and then the stat tests...
# we go back to the original data 
# without the summarise step
# use stat_summary() to calculate mean and plot the bars...
# and geom_jitter()
plot <- ggplot(tidy_data, aes(x=thy_status, y=value, fill = thy_status)) +
    stat_summary(fun="mean", geom="bar") +   # calculates means
    scale_fill_manual(values=c("blue", "red")) + 
    facet_wrap(~Cytokine, scales ="free_y") +
    geom_jitter() 

plot
```


With that done, we can now add the statistics


```{r add_stats_with_stat_compare_means}

# then the stats tests with stat_compare_means()
my_comparisons <- list( c("minus", "plus"))
plot + stat_compare_means(comparisons = my_comparisons)
# default Wilcoxon test - non parametric
# so values are different to the paper

```

The paper says that these are paired T-test so do again. 
Redo statistics with method = T.test. 

```{r add_t-tests_again}

# change method to t.test and paired - true
plot + stat_compare_means(comparisons = my_comparisons, method = "t.test",
        paired = TRUE, size = 2)
# these don't look the same as the values in the paper
# the paper says paired T-test but perhaps I don't have the pairings correct...
# I don't think there is a way to do this without more information
# from the authors... 

```


Replace the numbers with stars to indicate significance. 

```{r stars}

# can add some stars for significance but as the values aren't the same
# it won't be quite the same as the paper... 
# create an object, called plot, with the graph 
plot <- plot + stat_compare_means(comparisons = my_comparisons, method = "t.test",
        paired = TRUE, size = 2,
        symnum.args = list(cutpoints = c(0,0.0001,0.005,0.001, 1), 
            symbols = c("****","***", "**", "ns"))) +
    theme(strip.background = element_rect(fill = FALSE))

# show the object
plot

```



```{r add_titles}

# add titles and reformat them a bit
# change the formatting and move legend
plot + 
    labs(x = "Thy Status",
        y = "Cytokine levels", 
        title = "Figure 4a",
        subtitle = "Croft et al, Nature, 2019") +
    theme_classic() +
    theme(strip.background = element_rect(fill = "white", colour = "white")) +
    theme(legend.position="bottom")

```


pagebreak \pagebreak


Now, I have two Exercises for you.
These will test your understanding and really build on what you have learned. 

### Exercise 3
This exercise is to redraw these graphs as box and whisker plots and then as
violin plots. If you need to look back at the code from Module 1, that's OK. 
It's also OK to look up documentation using the help within R. 

### Answers for Exercise 3

The code here looks complex but most of it is just cut and pasted from 
above. 
If you have nice code, cut and pasting is just fine. 
This code is also on [Github]([Github here](https://github.com/brennanpincardiff/R4Biochemists201/blob/master/R/module_201_2_ans_ex_3.R))

```{r exercise_3}
# answers_201_2_exercise_3
my_comparisons <- list( c("minus", "plus"))

boxplot <- ggplot(tidy_data, aes(x=thy_status, y=value, fill = thy_status)) + 
    geom_boxplot() +
    scale_fill_manual(values=c("blue", "red")) +
    facet_wrap(~Cytokine, scales ="free_y") +
    geom_jitter() 

boxplot <- boxplot + stat_compare_means(comparisons = my_comparisons, method = "t.test",
                                  paired = TRUE, size = 2,
                                  symnum.args = list(cutpoints = c(0,0.0001,0.005,0.001, 1), 
                                                     symbols = c("****","***", "**", "ns"))) +
    theme(strip.background = element_rect(fill = FALSE))

boxplot + 
    labs(x = "Thy Status",
         y = "Cytokine levels", 
         title = "Figure 4a",
         subtitle = "Croft et al, Nature, 2019") +
    theme_classic() +
    theme(strip.background = element_rect(fill = "white", colour = "white")) +
    theme(legend.position="bottom")


v_plot <- ggplot(tidy_data, aes(x=thy_status, y=value, fill = thy_status)) + 
    geom_violin() +
    scale_fill_manual(values=c("blue", "red")) +
    facet_wrap(~Cytokine, scales ="free_y") +
    geom_jitter() 

v_plot <- v_plot + stat_compare_means(comparisons = my_comparisons, method = "t.test",
                                        paired = TRUE, size = 2,
                                        symnum.args = list(cutpoints = c(0,0.0001,0.005,0.001, 1), 
                                                           symbols = c("****","***", "**", "ns"))) +
    theme(strip.background = element_rect(fill = FALSE))

v_plot + 
    labs(x = "Thy Status",
         y = "Cytokine levels", 
         title = "Figure 4a",
         subtitle = "Croft et al, Nature, 2019") +
    theme_classic() +
    theme(strip.background = element_rect(fill = "white", colour = "white")) +
    theme(legend.position="bottom")

```


### Exercise 4
For exercise 4, I would like you to take the data that you tidied up for 
exercise 1, plot it and add statistics. I think you should be able to do that 
at this point. If you have questions, feel free to add. Answers to the code is 
below. 

### Answers for Exercise 4
This exercise took me quite a while. 
I included the addition of error bar which I didn't use in the example above. 
There is lots of code here so I have done a code demo video as well. 
Try to do as much as you can by yourself but don't worry if you don't get it
all first time around. 
Go through the code demo and the code and you will learn lots. 
This code is also on [Github]([Github here](https://github.com/brennanpincardiff/R4Biochemists201/blob/master/R/module_201_2_ans_ex_4.R)).


```{r exercise_4}

# read in easier data than the download.
link <- ("https://raw.githubusercontent.com/brennanpincardiff/RforBiochemists/master/data/fig_2d_altered.xlsx")
download.file(url=link, destfile="fig2d_altered.xlsx", mode="wb")

data <- read_excel("fig2d_altered.xlsx", skip = 1)

# gather(), separate() and mutate() into factors...
tidy_data_2 <- data %>%
    gather() %>%
    separate(key, into=c("names", "DTR_status"), sep = "/") %>%
    # need to turn the names into factors...
    mutate(DTR_status = factor(DTR_status)) %>%
    mutate(names = factor(names, c("bone_eros_CT", "bone_form_CT", 
                                   "area_bone_eros", "area_bone_form",
                                   "area_pan", "area_dest_cart")))


# simple bar plot calculating the means and using facet_wrap()
ggplot(tidy_data_2, aes(x=DTR_status, y=value, fill = DTR_status)) +
    stat_summary(fun="mean", geom="bar") +   # calculates means
    scale_fill_manual(values=c("white", "grey")) + 
    facet_wrap(~names, scales ="free_y", ncol = 2)

# to put on error bars need to summarise to calculate mean and SD. 
# more on error bars: http://www.sthda.com/english/wiki/ggplot2-error-bars-quick-start-guide-r-software-and-data-visualization
tidy_data_2 %>%
    group_by(names, DTR_status) %>%
    summarise(sd = sd(value), value = mean(value)) -> tidy_data_stats_2

# just stats first...
my_comparisons <- list( c("DTR-", "DTR+"))
p1 <- ggplot(data = tidy_data_2, aes(x=DTR_status, y=value)) + 
    stat_compare_means(comparisons = my_comparisons, method = "t.test",
                 paired = TRUE, size = 3,
                 symnum.args = list(cutpoints = c(0,0.0001,0.005,0.001, 1), 
                                   symbols = c("****","***", "**", "ns"))) +
    theme_classic2() +
    theme(strip.background = element_rect(fill = FALSE)) +
    facet_wrap(~names, scales ="free_y", ncol = 2)

p1
# to see the stars you need to Zoom... 

# then add bars
p1 <- p1 +  
    geom_bar(data = tidy_data_stats_2, 
             aes(x=DTR_status, y=value,  fill=DTR_status),
        stat="identity",
             position=position_dodge(),
        color = "black") +  
    scale_fill_manual(values=c("white", "grey"), name = "") +
    theme(legend.title = NULL, legend.position="top") +

    facet_wrap(~names, scales ="free_y", ncol = 2,              
               # moving title to side of plot
               strip.position = "left",
               # full titles for the y-axis
               labeller = as_labeller(c(bone_eros_CT = "Bone eroision \n score micro-CT",
                                        bone_form_CT= "Bone formation \n score micro-CT",
                                        area_bone_eros= "Area bone \n erosion (%)",
                                        area_bone_form = "Area bone \n formation (%)",
                                        area_pan = "Area pannus \n tissue (%)", 
                                        area_dest_cart = "Area destained \n cartilage (%)"))) +
    # geom_errorbar() adds the error bars :-)
    geom_errorbar(data = tidy_data_stats_2, 
                  aes(ymin=value-sd, ymax=value+sd), 
                  width=.2,
                  position=position_dodge(.9))

### then add geom_jitter to put points at the top...

p1 <- p1 + geom_jitter(data = tidy_data_2, 
                       aes(x=DTR_status, y=value))

# then add styling and titles...
p1

p1 <- p1 + ylab(NULL) + xlab(NULL) +
    # move strips outside of numbers
    theme(strip.background = element_blank(),
          strip.placement = "outside") + 
    # add labels...
    labs(title = "Figure 2d",
         subtitle = "Croft et al, Nature, 2019")

```
pagebreak \pagebreak
```{r exercise_4_large_print, fig.height=12, fig.width = 9}
p1
```



### Review what we have learned
* We have imported data from the internet.

* Our used our standard workflow
    + it was very messy so we changed it in Excel first
    + then we tidy the data with dplyr::pivot_longer(), separate() and mutate()
    + calculate means with dplyr::summarise()
    + visualise the data with ggplot2::geom_bar()
    + using facet_wrap() to make multiple plots
    
* Calculating some statistics
    + calculating statistics with t.test()
    + adding these to our plots with ggpubr::stat_compare_means()
    + by setting up comparisons and using method = "t.test"
    
* We have really deepened our learning of ggplot2
    + by revisiting geom_boxplot() and geom_violin() again
    + combining geoms with different data 
    + using the labeller argument in facet_wrap() 
    + using various theme() argements


### Resources
* You could move onto Module 3 -.
* The chapter of the ggplot2 book about facetting[https://ggplot2-book.org/facet.html]
* The site for ggpubr with worked examples [https://rpkgs.datanovia.com/ggpubr/]
* I like the (STHDA website)[http://www.sthda.com/english/wiki/ggplot2-essentials] - lots of worked examples
* I learned a lot about statistics from (Andy Field who has a website)[https://www.discoveringstatistics.com/].
