---
title: "R for Biochemists 201 Module 4 - Writing functions & Exploring Packages"
author: "Paul Brennan, Cardiff University, BrennanP@cardiff.ac.uk"
date: "Nov 2021"
output: 
# html_document
  word_document: default
#  pdf_document: default
---

### Welcome to Module 4 of the R for Biochemists 201 Training module.

In this module, we explore functions and packages. We will also talk
about version control with a focus on Github. 
In our pilot, learners took a about 2.5 hours to complete this module. 
Please don't worry if it takes you longer. There is a lot of material 
in this module. 
Rather, please be patient with yourself and try to keep learning. Take breaks
as you need and ask questions if you need help. 

There is text to read, video demonstrations,
example R Markdown documents and exercises. 
We hope you enjoy the module.

We welcome feedback and comments.

### How to use this module 
1. To learn from this module, cut and paste the R code into R-Studio and then run the code line by line. 
2. See if you can make the script work
3. Watch the demonstration videos to see how the code works
4. Look at how the 'Global Environment' changes.
5. Change the code and test it
6. Try the exercises
7. Extend your learning using the resources at the end

### Learning Objectives for Module 4
* Write a function 
* Document a function
* Introduce version control and Git
* Explore some packages 


As we deepen our knowledge and usage of R, it becomes useful to create 
functions. Functions can create a workflow, can be combination of other
functions or can be a new set of actions or calculations.

When are you ready to create a function? Well, if you find yourself using the
same commands or pieces of workflow multiple times then it suggests that you
could benefit from creating a function that brings these together. If you
generate data in the same way and analyse in the same way, you could benefit
from a workflow that brings together. 

Three is usually the 'magic' number. If you have or plan to repeat the same
analysis three times you could probably write better code and reduce your work
by writing some functions. 

To create a function, we use the function() function. 
Let's look at Croft Figure 4f.

### Loading the packages we need
```{r packages_we_need, message=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
```

pagebreak \pagebreak

### The Experiment
We are going to explore the data in 
[Figure 4f](https://www.nature.com/articles/s41586-019-1263-7/figures/4) 
from the Croft (2019) paper. There are panels about Leukocytes, Neutrophils and
Macrophages so we are going to create a function that links together the
workflow required for each of the three panels. 

The [data is available](https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-019-1263-7/MediaObjects/41586_2019_1263_MOESM10_ESM.xlsx).

The Excel file has 9 worksheets - each sheet corresponding to a part of 
Figure 4. 

Our workflow is:

1. Import the data
2. Tidy our data (also known as wrangling or munging)
3. Transform or summarise
4. Visualize




### Writing functions
##### First video - writing a function... 
```{r get_data_for_Fig_4f, message=FALSE}
link <- "https://static-content.springer.com/esm/art%3A10.1038%2Fs41586-019-1263-7/MediaObjects/41586_2019_1263_MOESM10_ESM.xlsx"

# the download.file() function downloads and saves the file with the name given
download.file(url=link, destfile="file.xlsx", mode="wb")

# check worksheet names 
sheet_names <- excel_sheets("file.xlsx")
data <- read_excel("file.xlsx", sheet_names[6])
```

As with previous data sets from this manuscript the raw data is quite Untidy. 
I have simplified it a little in Excel and it is available to download from
Github with this code.

```{r get_easier_data_for_Fig_4f}
link <- c("https://github.com/brennanpincardiff/R4Biochemists201/blob/master/data/fig_4f.xlsx?raw=true")

# the download.file() function downloads and saves the file with the name given
download.file(url=link, destfile="file.xlsx", mode="wb")

data <- read_excel("file.xlsx")
```


Our workflow is basically the same as before. 

Steps:

1. Extract cell type we want to plot - this will be an argument of our function
2. Transform the data
3. Generate the plot
4. Customise the plot

Usually, I try to make a script work for one part of the data and then turn 
that script into a function. Once it is a function, I test it on the starting
data and then use it on the rest of the data. 

This is the script for our workflow:
```{r first_workflow}
# select out the rows we want with filter() function for "Leucocytes"
# gather without cell_type and treatment
# mutate treatment into a factor to plot in the correct order
data %>% 
    filter(cell_type =="Leucocytes") %>% 
    gather(cell_number, mice, -cell_type, -treatment) %>%
    mutate(treatment = factor(treatment, c("Sham_1", "THY-", "Sham_2","THY+"))) -> data_2

# make the plot
set.seed(1)   # geom_jitter has a random element so set.seed() makes this same
plot_l <- ggplot(data = data_2,
    aes(treatment, mice, colour = treatment)) +
    geom_boxplot() +
    geom_jitter(width=0.15) +
    theme_bw() +
    labs(title = "Leucocytes", x="", y="",
         subtitle = "Croft et al (2019) Nature 570:246–251 (2019)") +
    theme_classic() + theme(legend.position="none")

plot_l
```

Now we turn it into a function using the function() function. 
In the brackets are the arguments for our function. 

```{r make_function}
## make a function called my_workflow

my_workflow <- function(data, a_cell_type){
    # select out the rows we want with filter() function
    data %>% 
        filter(cell_type == a_cell_type) %>% 
        gather(cell_number, mice, -cell_type, -treatment)  %>%    
        mutate(treatment = factor(treatment, c("Sham_1", "THY-", "Sham_2","THY+"))) -> data_2
        
    # make the plot
    set.seed(1)
    plot <- ggplot(data = data_2,
        aes(treatment, mice, colour = treatment)) +
        geom_boxplot() +
        geom_jitter(width=0.15) +
        theme_bw() +
        labs(title = a_cell_type, x="", y="",
             subtitle = "Croft et al (2019) Nature 570:246–251 (2019)") +
        theme_classic() + theme(legend.position="none")
    
    # this returns the plot that we want.   
    return(plot)
}

```

Please note here is that a new part has been created in my Global
Environment entitled Functions. 
In this new part of my Global Environment is the function called my_workflow.

```{r check_function}
# check_function
cell_types <- unique(data$cell_type)
cell_types[1] # this is Leucocytes and should work...
leuco_plot <- my_workflow(data, cell_types[1])
leuco_plot

# do a visual check that the same object as the script above
# it is possible to automatic checks too but that's for the future. 
```


Now use the function to generate two new plots with different cell types. 

```{r use_function}
# now plot for other cell types

cell_types[2] # this is Leucocytes and should work...
neutro_plot <- my_workflow(data, cell_types[2])
macro_plot <- my_workflow(data, cell_types[3])

# now we have three plots all made the same way.

neutro_plot
macro_plot
```


#### Exercise 1
Please create a function to add the statistical tests that we have used before. 
As I mentioned, I usually approach this by getting it right for one plot first
and then test and use the function. 
There are different ways to do this but do have a play and try to work out
what to do. 

#### Answers for Exercise 1
Here is an example of a function that will add the statistical tests.
```{r exercise_1_answer} 
# module_201_4_ans_ex_1

### activate the packages
library(tidyverse)
library(readxl)
library("ggpubr")

### download the data
link <- c("https://github.com/brennanpincardiff/R4Biochemists201/blob/master/data/fig_4f.xlsx?raw=true")

# the download.file() function downloads and saves the file with the name given
download.file(url=link, destfile="file.xlsx", mode="wb")

data <- read_excel("file.xlsx")

### this function is will add the ANOVA 
### is uses the ggboxplot() from the ggpubr package
stat_flow <- function(data, a_cell_type){
    # select out the rows we want with filter() function
    data %>% 
        filter(cell_type == a_cell_type) %>% 
        gather(cell_number, mice, -cell_type, -treatment)  %>%    
        mutate(treatment = factor(treatment, c("Sham_1", "THY-", "Sham_2","THY+"))) -> data_2
    
    ggboxplot(data_2, x = "treatment", y = "mice", add = "jitter",
              xlab = "",
              ylab = "") +
        stat_compare_means(method = "anova")
    }

### use the function
p1 <- stat_flow(data, "Leucocytes")
p2 <- stat_flow(data, "Neutrophils")
p3 <- stat_flow(data, "Macrophages")

### arrange the three graphs on one page...
figure <- ggarrange(p1, p2, p3, nrow=1,
          labels = c("A.Leucocytes",
                     "B.Neutrophils",
                     "C.Macrophages"))

# http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/

annotate_figure(figure, 
                top = text_grob("Figure 4f from Croft et al (2019) Nature 570:246–251",
                color = "red", face = "bold", size = 14))

# this figure looks best on R-Studio at a width of 1000 and a height of 600
```


### Our second advanced theme is documentation
##### Second video - document your function... 

We can create functions in our script files or R-markdown files. However, 
if we are using functions across multiple files, we might want to save them
separately.

To do this, create a new R script file with a meaningful name and cut and paste
our fuction. In this case, we are making a file called `workflow_fig4f.R`

To use the function, the source() function will bring it into the Global
Environment. 
```{r bring_in_function, eval=FALSE} 
source("/Users/paulbrennan/Documents/R4Biochemists201/R/workflow_fig4f.R")

```

To help other users or our future selves, we are going to document this
function. Documentation is key to your collaborators, remembering what
and why you have done something and if you 
ever want to share your functions or workflows with others. 
We could use hash tags but there are other ways to document functions in 
particular. These use sets of tags called Roxygen tags. 
There is a short cut that will do this Ctrl+Shift+Alt+R. 

The Roxygen tags use #' and @ symbols. Here is an example:

```{r roxygen_demo, eval=FALSE} 
#' Title
#'
#' @param data 
#' @param a_cell_type 
#'
#' @return
#' @export
#'
#' @examples

@param refers to paramaters 
@return tells you what is returned. 
@export is necessary if you are including your function in a package. 
```


We can add more information so that we know what the function does. 

### Exercise 2
Above I hope that you created your own function to add the statistical tests. 
Please you add the documentation to that now.

### Answers for Exercise 2
```{r roxygen_, eval=FALSE} 

#' A workflow to create graphs for Figure 4f and add an ANOVA
#' It uses the ggboxplot() from the ggpubr package
#'
#' @param data 
#' @param a_cell_type 
#'
#' @return a ggplot object created by ggboxplot
#'
#' @examples
#' 
### this function is will add the ANOVA 
### 
stat_flow <- function(data, a_cell_type){
    # select out the rows we want with filter() function
    data %>% 
        dplyr::filter(cell_type == a_cell_type) %>% 
        dplyr::gather(cell_number, mice, -cell_type, -treatment)  %>%    
        dplyr::mutate(treatment = factor(treatment, c("Sham_1",
                                                      "THY-", 
                                                      "Sham_2",
                                                      "THY+"))) -> data_2
    
    ggpubr::ggboxplot(data_2, x = "treatment", y = "mice", add = "jitter",
              xlab = "",
              ylab = "") +
        ggpubr::stat_compare_means(method = "anova")
    }
```

### Version control
##### Fourth video - version control on Github 
Version control is a key element of becoming a better programmer. 
It is a way of recording changes you have made to a script, to a function or
to a workflow. One of the most commonly used methods is called Git. 
One way to use it is through Github, a web version of Git. 

There is a video of how to use it and a task to do.

The steps:
* Create your own account on Github 
* Fork this Sandpit repository
* Make a change in your own fork - add something maybe your name and your favourite molecule.
* Commit the change with a comment
* Create a Pull Request for me to bring your code/text/link into the sandpit. 
* Repeat as many times as it takes for you to learn the process
* Don't forget to fetch and merge from upstream. This avoids conflicts. 

##### Fifth video - bringing together R-Studio and Githib

Next step shown is a video shows you how to get R-Studio working with Github, 
how to commit a change and push it online.

You will need to install Git onto your computer. Depending on your computer and
your technical ability, this may take a little practice. 

Please check out this 
[Happy Git for R link](https://happygitwithr.com/connect-intro.html) with good
information about doing this. You will need Git on your computer to integrate a
Github Repo with R-Studio and to allow you to make R-Studio work with
repositories on Github. 

This video shows you how to pull down a Github repo into R-Studio. 
On Github, please fork the 
[RforBiochemists201 repository](https://github.com/brennanpincardiff/R4Biochemists201)
so that you have your
own copy. Then please try create a project with your own RforBiochemists201
repo through R Studio. The RforBiochemists201 repo has more files in it now
than it did when I recorded this video. You may be able to make and COMMIT
changes and PUSH them. However, this requires you to link up R-Studio to Github
which will probably require the creation of Personal access tokens. For
information on this, please 
[check this link](https://happygitwithr.com/https-pat.html).



### Exercise 3
If you don't already have one, you will need to create an account
for yourself on [Github.com](https://github.com/).

Then, try creating your own My_Learning_R prjoect on Github. 

* Put in a Readme. 
* Clone it into a new R Studio project. 
* Write some code for the analysis of some data. 
* Stage your change and commit the change 
* Push those changes onto Github. 
* See if you can make it all work together. 
* Make more changes and repeat the process. 
* Share the address of your Github repo to share your progress. 

If you can make this work, you have taken a key next step towards being a
programmer. Do ask questions if you need help. 

pagebreak \pagebreak

### Packages
##### Fourth video - packages 

Packages are group of functions that are available to download and explore. 
There are lots of packages available that you might want to play with. 
You might want to adapt these to your own workflows, change functions
and even contribute to them as your skills develop. 

Sources of packages:

1. [CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html)
2. [Bioconductor](http://bioconductor.org/)
3. [R-OpenSci](https://ropensci.org/)
4. [Github](https://github.com/)

A good place to explore for packages is Bioconductor. These are
packages for Bioinformatics and can be used to explore biological data.

For example, perhaps you want to look at some flow data. 
There is a package for that on Bioconductor. 
You need to install BiocManager and then install the flowCore package. Here
is a demonstration of this using some of my data. 

```{r download_from_Bioconductor}
# install BiocManager
# install.packages("BiocManager")
# BiocManager::install("flowCore")

library("flowCore")
library(ggplot2)
library(ggpubr)


link1 <- "https://github.com/brennanpincardiff/R4Biochemists201/blob/master/data/cfse_data_20111028_Bay_d7/A01.fcs?raw=true"

# download first data file
download_data <- function(link){
    download.file(url=link, destfile="file.fcs", mode="wb")
    data <- flowCore::read.FCS("file.fcs", alter.names = TRUE)}

data <- download_data(link1)

#with colours indicating density
colfunc <- colorRampPalette(c("white", "lightblue", "green", "yellow", "red"))
# this colour palette can be changed to your taste 

vals <- as.data.frame(exprs(data))
ggplot(vals, aes(x=FSC.A, y=SSC.A)) +
    ylim(0, 500000) +
    xlim(0,5000000) +
    stat_density2d(geom="tile", aes(fill = ..density..), contour = FALSE) +
    scale_fill_gradientn(colours=colfunc(400)) + # gives the colour plot
    geom_density2d(colour="black", bins=5) # draws the lines inside


# write function for plotting CFSE data...
plot_cfse <- function(data){
    # this exprs() function pulls out the numbers. 
    vals_1 <- as.data.frame(exprs(data))
    # I want to exclude small debris and just analyse cells. 
    vals_f <- dplyr::filter(vals_1, FSC.A>1000000)
    # then draw the plot
    ggplot(vals_f, aes(x=FL1.H)) + geom_density() +
        xlim(1000,1000000) + scale_x_log10()
}

# plot first data set
p1 <- plot_cfse(data)

# download and plot second data set
link2 <- "https://github.com/brennanpincardiff/R4Biochemists201/blob/master/data/cfse_data_20111028_Bay_d7/A02_top_dose.fcs?raw=true"
data2 <- download_data(link2)
p2 <- plot_cfse(data2)

# download and plot third data set
link3 <- "https://github.com/brennanpincardiff/R4Biochemists201/blob/master/data/cfse_data_20111028_Bay_d7/A04.fcs?raw=true"
data3 <- download_data(link3)
p3 <- plot_cfse(data3)


text <- paste("Each decrease in fluorescence represents a population division.")

text.p <- ggparagraph(text = text, face = "italic", size = 11, color = "black")

ggarrange(p1, p2, p3, text.p, ncol=1,
          labels = c("A.Cell proliferation in control",
                     "B.Inhibited by Bay top dose",
                     "C.Half Bay dose inhibits less"))

```


You can download and use a package from Github. 
Explore gganatogram as an example. 
```{r download_from_github}
## install from Github
# remove the hash tag to run. 
#devtools::install_github("jespermaag/gganatogram")
library(gganatogram)
library(dplyr)
library(viridis)
library(gridExtra)
 
organPlot <- data.frame(organ = c("heart", "leukocyte", "nerve", "brain", "liver", "stomach", "colon"), 
 type = c("circulation", "circulation",  "nervous system", "nervous system", "digestion", "digestion", "digestion"), 
 colour = c("red", "red", "purple", "purple", "orange", "orange", "orange"), 
 value = c(10, 5, 1, 8, 2, 5, 5), 
 stringsAsFactors=F)

 head(organPlot)

gganatogram(data=organPlot, fillOutline='#a6bddb', organism='human', sex='female', fill="colour")

```


### Exercise 4
Download and draw a brain plot using 
the ggseg package (https://github.com/LCBC-UiO/ggseg)


### Answers Exercise 4
Check out this Vignette as well maybe
(https://cran.r-project.org/web/packages/ggseg/vignettes/ggseg.html)
```{r exercise_4_download_ggseg}
## install from CRAN
# remove the hash tag to run. 
# install.packages("ggseg")
library(ggseg)
library(ggplot2)
ggplot() +
    geom_brain(atlas = dk) +
    theme(legend.position = "none")

# this also gives a nice picture...
ggseg(atlas=aseg)
```

### Looking at packages on Github
It is interesting to check out the details of a package on Github if it
is there. Here is a the Github repo for
[gganatogram](https://github.com/jespermaag/gganatogram). By looking at various
packages you can get a feeling for how they
are organised. Here is a little about how packages are structured. 

Basic anatomy of a package on Github...

* Readme
* Build badge - passing is good.
* R folder - has the code - you can look at this and adapt.
* man folder - has the documentation - it is generated by the Roxygen tags.
* test folder - good packages test their code - the tests are here.
* Vignettes - some example code that you can explore. 


You can also download all the material from a package as a separate RStudio
project. This will fork the package and you can play, edit, adapt and make
changes if you like. 

There is also a very good
book about [Packages](http://r-pkgs.had.co.nz/) and how they can be organised. 

### Exercise 5
Try to complete the following

* Within Github, fork the drawProteins repository to your own area. The original is here: https://github.com/brennanpincardiff/drawProteins.
* Then, using R-Studio, create a new project and download your own drawProteins package.
* Open the vignette - an R-Markdown file showing how to use drawProteins. 
* Try Knitting the R-Markdown file. 
* Create your own R-Markdown file to use drawProteins to generate a schematic 
of your favourite protein (or adapt your file from Module 3). 
* Try publishing your R-Markdown on Rpubs perhaps. Share the link online. 
* Remember you can use version control to store your work as you go along. 


Check out the online 
[vignette for drawProteins](https://www.bioconductor.org/packages/release/bioc/vignettes/drawProteins/inst/doc/drawProteins_BiocStyle.html). 
Remember you can show some or
all of the of the code in your RMarkdown file. 

This exercise should integrate lots of what you have learned in Module 3 and 4. 
I haven't supplied a solution to this. I would rather see your results on RPubs,
Github or both. 

Please ask questions if you need help. 


### Answers Exercise 5
```{r download_drawProteins, eval=FALSE}
# download drawProteins from Bioconductor
# install BiocManager
# install.packages("BiocManager")
# BiocManager::install("drawProteins")

library(drawProteins)
# show the vignette for drawProteins in the Help tab
vignette("drawProteins_BiocStyle")
# sample code is in the vignette... explore it. 

# integrate your learning by adding the code and output to your RMarkdown file
# and/or the Github page you created above...
```

Here is a little code and the output to illustrate.

```{r exercise_5_drawProteins_example}
library(drawProteins)
drawProteins::get_features("Q04206") -> rel_json
feature_to_dataframe(rel_json) -> rel_data
draw_canvas(rel_data) -> p
p <- draw_chains(p, rel_data)
p <- draw_domains(p, rel_data)
p <- draw_regions(p, rel_data)
p + theme(legend.position="bottom") 
```



pagebreak \pagebreak

### Review what we have learned
* We created a function
* We documented our function using Roxygen tags
* We learned the basics of version control using Github
* We looked and explored packages on Bioconductor and Github

### Resources
* [Software Carpentry piece about writing functions](https://swcarpentry.github.io/r-novice-inflammation/02-func-R/) 
* [Object documentation](https://r-pkgs.org/man.html) - chapter on documentation
* [Happy Git and Github for the useR](https://happygitwithr.com/) - helpful advice for using Git with R. 
* [Bioconductor](http://bioconductor.org/) - lots of bioinformatics resources for R. 
* [R Open Sci](https://ropensci.org/) - a great resource for open and reproducible science
* [R Packages](https://r-pkgs.org/) - the definitive book on packages
